---
title: "HW5"
author: "Raja Prabakaran"
date: "2024-06-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(reshape2)
library(plotly)
library(caret)
library(factoextra)
library(rpart)
library(pROC)
```

# a. Data gathering and integration

```{r}
bundesliga <- read.csv("Bundesliga_player.csv")
head(bundesliga)

```

# b. Data Exploration

##Summary Statistics

```{r}

summary(bundesliga)

```

##Distributions of Key Variables

###Combined Histogram for Numerical Variables

```{r}
numerical_vars <- c("age", "height", "price", "max_price")

df_long <- melt(bundesliga, measure.vars = numerical_vars)

combined_histogram <- ggplot(df_long, aes(x = value, fill = variable)) +
  geom_histogram(binwidth = 5, color = "black", position = "identity", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free") +
  labs(x = "Value", y = "Frequency", fill = "Variable") +
  ggtitle("Combined Histogram of Numerical Variables") +
  theme_minimal()

print(combined_histogram)

```

##Box Plots for Numerical Variables by Position

```{r}
bundesliga$position <- as.factor(bundesliga$position)

plots <- lapply(numerical_vars, function(var) {
  plot_ly(data = bundesliga, x = bundesliga[[var]], y = bundesliga$position, type = "box",
          color = bundesliga$position, colors = "Set3",
          marker = list(line = list(color = 'rgb(0,0,0)', width = 1)),
          boxmean = TRUE) %>%
    layout(title = paste(var, "vs Position"),
           xaxis = list(title = var),
           yaxis = list(title = "Position"),
           template = "plotly_white")
})


plots

```

# c. Data Cleaning



##Fixing Typos and Cleaning Strings

```{r}
bundesliga$nationality <- gsub("  ", "", bundesliga$nationality)
bundesliga$place_of_birth <- gsub("  ", "", bundesliga$place_of_birth)

```

##Changing Date Columns to Date Format

```{r}

bundesliga$contract_expires <- as.Date(bundesliga$contract_expires, format="%d/%m/%Y")
bundesliga$joined_club <- as.Date(bundesliga$joined_club, format="%d/%m/%Y")

```

##Removing Unnecessary Columns

```{r}
bundesliga <- bundesliga %>% select(-full_name, -player_agent,-place_of_birth, -contract_expires, -joined_club)
```


##Removing Missing Values

```{r}

bundesliga <- na.omit(bundesliga)

```
##Checking for Missing Values and Summary Statistics After Cleaning

###Cleaning

```{r}
summary(bundesliga)

```

###Visualizing Clean Data

```{r}
ggplot(bundesliga, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Age Distribution of Players (Cleaned Data)", x = "Age", y = "Count")

```

# d. Data Preprocessing

##Normalization

```{r}
bundesliga[numerical_vars] <- scale(bundesliga[numerical_vars])

```

##Creating Dummy Variables for Categorical Columns

```{r}
categorical_cols <- c('position', 'nationality', 'club')

bundesliga <- bundesliga %>%
  mutate(across(all_of(categorical_cols), as.factor)) %>%
  model.matrix(~.-1, data=.) %>%
  as.data.frame()

```

##Binning 'Age' into Categories: 'young', 'mid', 'old'

```{r}
bundesliga$age_group <- cut(bundesliga$age, breaks = 3, labels = c("young", "mid", "old"))

```




##Visualizing Preprocessed Data

```{r}
ggplot(bundesliga, aes(x = age_group)) +
  geom_bar(fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Age Groups (Preprocessed Data)", x = "Age Group", y = "Count")


```



# e. Clustering

##PCA and Clustering

```{r}


numeric_cols <- sapply(bundesliga, is.numeric)
bundesliga_numeric <- bundesliga[, numeric_cols]

bundesliga.pca <- prcomp(bundesliga_numeric, center = TRUE, scale. = TRUE)
summary(bundesliga.pca)
screeplot(bundesliga.pca, type = "l") + title(xlab = "PCs")

bundesliga_pca_df <- as.data.frame(bundesliga.pca$x)


```

## Hierarchical Clustering

```{r}
dist_mat <- dist(bundesliga_pca_df, method = 'manhattan')
hfit <- hclust(dist_mat, method = 'average')
plot(hfit)

fviz_nbclust(bundesliga_pca_df, FUN = hcut, method = "wss")
fviz_nbclust(bundesliga_pca_df, FUN = hcut, method = "silhouette")

h3 <- cutree(hfit, k = 3) 
fviz_cluster(list(data = bundesliga_pca_df, cluster = h3))

bundesliga_pca_df$Clusters <- as.factor(h3)

ggplot(data = bundesliga_pca_df, aes(x = PC1, y = PC2, col = Clusters)) +
  geom_point()

```

## K-Means Clustering

```{r}
preproc <- preProcess(bundesliga_pca_df, method = c("center", "scale"))
bundesliga_normalized <- predict(preproc, bundesliga_pca_df)

numeric_cols <- sapply(bundesliga_normalized, is.numeric)
bundesliga_normalized_numeric <- bundesliga_normalized[, numeric_cols]

fviz_nbclust(bundesliga_normalized_numeric, kmeans, method = "wss")
fviz_nbclust(bundesliga_normalized_numeric, kmeans, method = "silhouette")

fit_kmeans <- kmeans(bundesliga_normalized_numeric, centers = 2, nstart = 25) 

fviz_cluster(fit_kmeans, data = bundesliga_normalized_numeric)


```



# f. Classification

##Creating the Target Variable and Splitting the Data

```{r}
set.seed(123)

bundesliga$age_group <- as.factor(bundesliga$age_group)

index <- createDataPartition(y = bundesliga$age_group, p = 0.7, list = FALSE)
train_data <- bundesliga[index,]
test_data <- bundesliga[-index,]

predictors <- sapply(train_data, is.numeric) | sapply(train_data, is.factor)
train_data <- train_data[, predictors]
test_data <- test_data[, predictors]

names(train_data) <- make.names(names(train_data))
names(test_data) <- make.names(names(test_data))

```

##Training the k-NN Model

```{r}
train_control <- trainControl(method = "cv", number = 10)

knn_model <- train(age_group ~ ., data = train_data, method = "knn", trControl = train_control, tuneLength = 20)

print(knn_model)


```



```{r}
tree_model <- train(age_group ~ ., data = train_data, method = "rpart", trControl = train_control)

print(tree_model)


```



# g. Evaluation

#Confusion Matrix

```{r}
train_data$age_binary <- ifelse(train_data$age_group == "young", "young", "not_young")
test_data$age_binary <- ifelse(test_data$age_group == "young", "young", "not_young")

train_data$age_binary <- factor(train_data$age_binary, levels = c("young", "not_young"))
test_data$age_binary <- factor(test_data$age_binary, levels = c("young", "not_young"))

tree_model_binary <- train(age_binary ~ ., data = train_data, method = "rpart", trControl = train_control)

tree_predictions_binary <- predict(tree_model_binary, newdata = test_data)

tree_predictions_binary <- factor(tree_predictions_binary, levels = c("young", "not_young"))

tree_conf_matrix_binary <- confusionMatrix(tree_predictions_binary, test_data$age_binary)
print(tree_conf_matrix_binary)

```
##Precision and Recal


```{r}

cm_values <- as.numeric(tree_conf_matrix_binary$table)
true_negative <- cm_values[1]
false_positive <- cm_values[2]
false_negative <- cm_values[3]
true_positive <- cm_values[4]

precision <- true_positive / (true_positive + false_positive)
recall <- true_positive / (true_positive + false_negative)

precision
recall

```

##ROC Curve

```{r}
tree_pred_prob_binary <- predict(tree_model_binary, newdata = test_data, type = "prob")
head(tree_pred_prob_binary)

roc_tree_binary <- roc(response = test_data$age_binary, predictor = tree_pred_prob_binary[, "young"])

plot(roc_tree_binary, main = "ROC Curve for Decision Tree (Binary)")
legend("bottomright", legend = c("AUC: 1.0000"), lwd = 2)


```



## h. Report

- Refer pdf

## i. Reflection

- Refer pdf



